Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-14 paper.qmd
rewrote intro to uncert vis

# What is an Uncertainty Visualisation?
The lack of a consistent definition of uncertainty means that every statement that builds upon "uncertainty" will be ill-formed. This includes statements about uncertainty visualisation. This appears in the literature as a mountain of seemingly conflicting statements. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap __*). 
Cite Gap: Confusion about what "uncertainty visualisation" is
- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.

Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem. For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue. @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy. What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.

## Common misconceptions about uncertainty visualisation

### Defining "uncertainty visualisation" as one word

@@

### Misunderstanding the role of contextual information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

@@

>>>>

# What is Uncertainty Visualisation?
If we simply built the idea of uncertainty visualisation up from "uncertainty" which we already know has a precarious definition we would understand the field of "uncertainty" visualisation must also be precarious. You cannot quantify something that is not defined, and you can't visualise something you cannot quantify. If we cannot quantify the full concept uncertainty in any meaningful way (other than with examples such as missing values or a PDF) then there is no such thing as an "uncertainty" visualisation. Unfortunately this problem is not quite that simple.

A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words). When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap __*). @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.

These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a conerstone of some clearly questionable results within this field.

## Taxonomies of uncertainty visualisation

@@

## Contextual Information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

@@

## Fixes within "uncertainty visualisation"
Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem. For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue. @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy. What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.
